General purpose gene annotation using deep learning

### 2020-01-09 paper/main.tex
data split and selection

The selection of training genomes requires balancing multiple conflicting tradeoffs. On the one hand, we do want to train with as much data as possible, but not all data has a good enough quality to enable a powerful generalization. It is also very desirable to have a broad spread in phylogenetic distance as well as genome size and average gene length as the model is increasingly unlikely to generalize well beyond the borders given by the training data. It was also important to us to get experimental results within a couple of days as we tested many different data and model configurations, which limited the overall size of the training set and the model.

>>>>

The selection of training genomes requires balancing multiple conflicting tradeoffs. On the one hand, we do want to train with as much data as possible, but not all data has a good enough quality to enable a powerful generalization. It is also very desirable to have a broad spread in phylogenetic distance as well as genome size and average gene length as the model is increasingly unlikely to generalize well beyond the borders given by the training data. However, if the genomic patterns inside the data are too different from each other, it may be difficult for the model to generalize as well. It was also important to us to get experimental results within a couple of days as we tested many different data and model configurations, which limited the overall size of the training set and the model.

We employed an iterative approach to effectively select a proper set of training genomes. We first started with 3-4 genomes that were almost certain to be of the highest quality. We then evaluated the performance of the resulting model on all genomes individually (including the training genomes) and looked for candidates to add to the training set or to remove from it. This process was repeated multiple times concurrently with the model search until we saw no more room for substantial improvement given our computational constraints.

Once a set of training genomes was selected, we generated a validation set from it by splitting of 20\% of the FASTA sequences above and below the N90 of each genome, ensuring a proper distribution of large and short sequences in both the training and validation set. The validation set was only used to roughly judge the performance of our models during training. Evaluation of the annotations on all training and validation genomes was done regularly, as we ultimately only cared about the performance on fully unknown genomes.
